Make interactive loan word maps

### 2025-05-06 analysis/paper/loan word maps.qmd
tidying repo

---
title: "Make maps of loan words"
author:
  - Henning Schreiber:
      correspondence: "yes"
      email: janedoe@fosg.org
      orcid: 0000-0003-1689-0557
      institute: uhh

  - Nikolas Gestrich:
      institute: fi
      orcid: 0000-0003-1689-0558

institute:
  - uhh:
      name: Universität Hamburg
      address: 23 Science Street, Eureka, Mississippi, USA
  - fi: Frobenius Institute
title-block-published: "Last updated"  
date: now
date-format: long
format: 
  docx:
    reference-doc: "../templates/template.docx" # Insert path for the DOCX file
execute:
  echo: true
  warning: false
  message: false
  comment: "#>"
  fig-path: "../figures/"
  fig-dpi: 600
filters:
  - ../templates/scholarly-metadata.lua
  - ../templates/author-info-blocks.lua
  - ../templates/pagebreak.lua
bibliography: references.bib
csl: "../templates/journal-of-archaeological-science.csl" # Insert path for the bib-style
abstract: |
  Text of abstract
keywords: |
  keyword 1; keyword 2; keyword 3
highlights: |
  These are the highlights. 
prefer-html: true
---

@@

Next, we read in the data and join the entries to the glottolog entries. Modify the list of .csv files to suit your interests. If you want to upload new files, place these in the directory /analysis/data/raw_data.Make sure any new data has a glottolog language code in a varaible named "glott_code"

@@

# make sf geometry
test <- point_list[[1]] |> sf::st_as_sf(coords = c("lon","lat"))

# set crs
sf::st_crs(test) <- 4326
```

@@

#add point layer
add_sf_to_leaflet_map(test, label_column = "site")
```

## Make static maps (for printing etc)

Function to make static plots

```{r}
create_static_map <- function(data_sf, form_column = "form", style_column = "style", plot_title = NULL) {

  # Define the color palette for the "style" column, ensuring darker colors
  num_categories <- length(unique(data_sf[[style_column]]))
  dark_palette <- scales::brewer_pal(palette = "Dark2")(num_categories)

  # Create the base map with background
  base_map <- ggplot2::ggplot() +
    ggspatial::annotation_map_tile(tiles = "http://server.arcgisonline.com/ArcGIS/rest/services/Canvas/World_Light_Gray_Base/MapServer/tile/${z}/${y}/${x}") +
    ggplot2::geom_sf(data = data_sf, ggplot2::aes(color = !!rlang::sym(style_column)), show.legend = FALSE) +
    ggplot2::scale_color_manual(values = dark_palette) +
    ggplot2::theme_minimal() +
    ggplot2::theme(legend.position = "bottom") +
    ggplot2::labs(color = style_column)

  # Add the plot title if provided
  if (!is.null(plot_title)) {
    base_map <- base_map + ggplot2::ggtitle(plot_title)
  }

  # Crop the extent of the map to the data layers' extent
  base_map <- base_map + ggplot2::coord_sf(xlim = sf::st_bbox(data_sf)[c("xmin", "xmax")],
                                           ylim = sf::st_bbox(data_sf)[c("ymin", "ymax")])

  # Plot the data as strings contained in the form_column variable
  map_plot <- base_map +
    ggplot2::geom_text(data = data_sf, ggplot2::aes(x = sf::st_coordinates(data_sf)[,1],
                                                    y = sf::st_coordinates(data_sf)[,2],
                                                    label = !!rlang::sym(form_column),
                                                    color = !!rlang::sym(style_column)),
                       position = ggplot2::position_jitter(width = 0.1, height = 0.1), size = 3)

  # Return the ggplot object
  return(map_plot)
}
```

Make static plots

```{r}
create_static_map(data_list[[2]], plot_title = "fonio")
```

```{r}
create_static_map_tmap <- function(data_sf, form_column = "form", style_column = "style", plot_title = NULL) {

  # Define the color palette for the "style" column, ensuring darker colors
  num_categories <- length(unique(data_sf[[style_column]]))
  dark_palette <- scales::brewer_pal(palette = "Dark2")(num_categories)

  # Increase the boundaries of the plot
  bbox <- sf::st_bbox(data_sf)
  x_range <- bbox["xmax"] - bbox["xmin"]
  y_range <- bbox["ymax"] - bbox["ymin"]
  bbox["xmin"] <- bbox["xmin"] - 0.5 * x_range
  bbox["xmax"] <- bbox["xmax"] + 0.5 * x_range
  bbox["ymin"] <- bbox["ymin"] - 0.5 * y_range
  bbox["ymax"] <- bbox["ymax"] + 0.5 * y_range
  attr(bbox, "class") <- "bbox"

  # Get the background map using maptiles package without cropping
  background_map <- maptiles::get_tiles(data_sf, provider = "Esri.WorldGrayCanvas", crop = FALSE)

  # Create the tmap object
  tmap::tmap_mode("plot")
  map_plot <- tmap::tm_shape(background_map, bbox = bbox) +
    tmap::tm_rgb() +
    tmap::tm_shape(data_sf) +
    tmap::tm_text(text = form_column, col = style_column, 
                  col.scale = tmap::tm_scale(values = dark_palette)) +
    tmap::tm_layout(title = plot_title, legend.show = FALSE, bbox = bbox)

  # Return the tmap object
  return(map_plot)
}
```

```{r}
create_static_map_tmap(data_list[[1]])
```

## Make a network

Our network will be made using languages as nodes, which come with locations. We will then use the data in the list to create edges between the nodes.

```{r}
# Load necessary libraries
library(dplyr)
library(tidyr)

# Function to create edge list for a dataframe
create_edge_list <- function(df, df_name) {
  # Clean and transform the data
  df <- df |> 
    select(glott_code, style) |> 
    as_tibble()

  # Create edge list based on first letter of style
  edge_list_first_letter <- df |>
    mutate(first_letter = substr(style, 1, 1)) |>
    inner_join(., ., by = "first_letter") |>
    filter(glott_code.x != glott_code.y) |>
    select(glott_code.x, glott_code.y) |>
    distinct() |>
    mutate(type = df_name)

  # Create edge list based on first three characters of style
  edge_list_first_three <- df |>
    mutate(first_three = substr(style, 1, 3)) |>
    inner_join(., ., by = "first_three") |>
    filter(glott_code.x != glott_code.y) |>
    select(glott_code.x, glott_code.y) |>
    distinct() |>
    mutate(type = df_name)

  # Combine both edge lists
  combined_edge_list <- bind_rows(edge_list_first_letter, edge_list_first_three) |>
    distinct()

  return(combined_edge_list)
}

# Apply function to all dataframes in the list and combine results
all_edge_lists <- bind_rows(
  lapply(names(data_list), function(df_name) {
    create_edge_list(data_list[[df_name]], df_name)
  })
)

```

```{r}
# Load necessary libraries
library(dplyr)
library(tidyr)

# Subset the data
data_list_2 <- purrr::map(data_list, ~ {
  # Select the columns glott_code and style and convert to tibble
selected_df <- .x %>%
    as_tibble() %>%   # Convert to tibble
    select(glott_code, style) %>% # Select the relevant columns
    drop_na(style)

  return(selected_df)
})
```

```{r}

# Function to create edge list for a dataframe
create_edge_list <- function(df, df_name) {
  # Create edge list based on first letter of style
  df_first_letter <- df |>
    mutate(first_letter = substr(style, 1, 1))
  edge_list_first_letter <- df_first_letter |>
    inner_join(df_first_letter, by = "first_letter") |>
    filter(glott_code.x != glott_code.y) |>
    select(glott_code.x, glott_code.y) |>
    distinct() |>
    mutate(type = df_name, weight = 1)

  # Create edge list based on first three characters of style
  df_first_three <- df |>
    mutate(first_three = substr(style, 1, 3))
  edge_list_first_three <- df_first_three |>
    inner_join(df_first_three, by = "first_three") |>
    filter(glott_code.x != glott_code.y) |>
    select(glott_code.x, glott_code.y) |>
    distinct() |>
    mutate(type = df_name, weight = 1)

  # Combine both edge lists
  combined_edge_list <- bind_rows(edge_list_first_letter, edge_list_first_three) |>
    distinct()

  return(combined_edge_list)
}

# Apply function to all dataframes in the list and combine results
all_edge_lists <- bind_rows(
  lapply(names(data_list_2), function(df_name) {
    create_edge_list(data_list_2[[df_name]], df_name)
  })
)

# Summarize duplicates and calculate the weight
all_edge_lists_summarized <- all_edge_lists |>
  group_by(glott_code.x, glott_code.y, type) |>
  summarize(weight = sum(weight), .groups = 'drop')
```

Now, add the edge list and the node list to a tidygraph network

```{r}
library(tidygraph)

g <- tbl_graph(edges = all_edge_lists_summarized, directed = FALSE)

plot(g)
```

>>>>

---
title: "Make interactive loan word maps"
author:
  - Henning Schreiber:
      orcid: 0000-0001-9576-8036
      institute: uhh

  - Nikolas Gestrich:
      correspondence: "yes"
      email: gestrich@uni-frankfurt.de
      institute: fi
      orcid: 0000-0001-6376-1972

institute:
  - uhh:
      name: Asien-Afrika-Institut, Universität Hamburg
      address: Mittelweg 177, 20148 Hamburg
  - fi: 
      name: Frobenius Institute
      address: Norbert-Wollheim-Platz 1, 60323 Frankfurt am Main, Gemany
title-block-published: "Last updated"  
date: now
date-format: long
format: 
  docx:
    reference-doc: "../templates/template.docx" # Insert path for the DOCX file
execute:
  echo: true
  warning: false
  message: false
  comment: "#>"
  fig-path: "../figures/"
  fig-dpi: 600
filters:
  - ../templates/scholarly-metadata.lua
  - ../templates/author-info-blocks.lua
  - ../templates/pagebreak.lua
bibliography: references.bib
csl: "../templates/journal-of-archaeological-science.csl" # Insert path for the bib-style
abstract: |
  Text of abstract
keywords: |
  keyword 1; keyword 2; keyword 3
highlights: |
  These are the highlights. 
prefer-html: true
---

@@

Next, we read in the data and join the entries to the glottolog entries. Modify the list of .csv files to suit your interests. If you want to upload new files, place these in the directory /analysis/data/raw_data.Make sure any new data has a glottolog language code in a variable named "glott_code"

@@

# make sf geometry
points_sf <- point_list[[1]] |> sf::st_as_sf(coords = c("lon","lat"))

# set crs
sf::st_crs(points_sf) <- 4326
```

@@

#add point layer
add_sf_to_leaflet_map(points_sf, label_column = "site")
```
